```{r setup, include = F}
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@
# @@@@@ Knitr Options
# @@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@@

# Set root directory to the project directory
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())


# Set default knitr options: 
# Suppress warnings and messages, cache chunks, 
#  set default figure size to 6x8 at 300 dpi, and save a png and pdf
knitr::opts_chunk$set(warning = F, message = F, collapse = T, cache = T,
    fig.height = 6, fig.width = 8, dpi = 300, # 6x8" @ 300dpi:1800x2400=4.3MP
    dev = c('png', 'pdf'), dev.args = list(pdf = list(onefile = F)))

```

overdispersion test
```{r}
overdispersion_test <- function(model, type = "pearson"){
    
    # Get the pearson residuals
    residuals <- resid(model, type = type)
    
    # Get the residual degrees of freedom of the model
    df <- df.residual(model)
    
    # Sum of residual deviance
    dev <- sum(residuals ^ 2)
    
    # Overdispersion = sum of squared residuals / residual degrees of freedom
    ratio <- round(dev / df, 3)
    
    # P-value 
    pvalue <- round(pchisq(dev, df, lower.tail = FALSE), 3)
    
    # Get the formula
    f = paste(as.character(formula(model))[2:3], collapse = " ~ ")
    
    # Get the model name
    name <- deparse(substitute(model))
    cat("Overdispersion ratio for model:", name, "\nformula:", f, 
        "\n\nAcceptable range: 1 - 1.4\nOverdispersion ratio:",
        ratio, " df:", df, " p =", pvalue, "\n", 
        ifelse(pvalue < 0.05, "Data are OVERDISPERSED\n", 
        "Data are NOT OVERDISPERSED\n"))
    # Return all the parameters
    return(c(ratio = ratio, deviance = dev, df = df, pvalue = pvalue))
    
}
```



loading in packages
```{r}
library(tidyverse)
library(ggplot2)
library(readr)
library(metafor)
library(sjPlot)
library(lme4)
library(car)
library(emmeans)
library(lubridate)
library(glmmTMB)

# added these libraries:
library(brms)
library(tidybayes)
library(viridis)

```
loading in data

```{r}

data <- read.csv("Data/cubee2026.csv")


```
subtracting mortality, adding day column, and obs 
```{r data_wrangling, echo = F}
dat <- data %>%
  mutate(
    Date = mdy(Date),
    NDB = as.numeric(NDB),
    TotalM = as.numeric(TotalM),
    Treatment = as.factor(Treatment)
  ) %>%
  arrange(CageID, Date) %>%
  group_by(CageID) %>%
  mutate(
    CageID = factor(CageID),
    TotalM_cage = first(na.omit(TotalM)),
    TotalD = cumsum(replace_na(NDB, 0)),
    Alive = TotalM_cage - TotalD,
    Day = as.numeric(Date - min(Date, na.rm = TRUE)),
   Risk = as.numeric(Alive + NDB),
   logRisk = log(Risk)) %>%
  filter(!is.na(logRisk)) %>%
  ungroup() %>%
    mutate(obs_ID = factor(dplyr::row_number()),
           NBE = as.numeric(NBE),
           cens_count = case_when(
             Day == 14 ~ NBE + 10,
             Day == 21 ~ NBE + 10, 
             TRUE ~ NBE
           )) %>%
  select(-TotalM_cage)

```
models
```{r}

#cumulative survival
mod1 <-  glmmTMB(cbind(Alive, TotalD) ~ Treatment * Day +
                  (1 | CageID),
            family = binomial(link = "logit"), data = dat) 

overdispersion_test(mod1) 
plot_model(mod1, type = "pred",
           terms = c("Day", "Treatment"))

mod2 <-  glmmTMB(cbind(Alive, TotalD) ~ Treatment* Day +
                  (1 | obs_ID),
            family = binomial(), data = dat) 

overdispersion_test(mod2)

mod3 <- glmmTMB(cbind(Alive, TotalD) ~ Treatment + Day +
                  (1 | CageID) + (1 | obs_ID),
            family = binomial(), data = dat) 

overdispersion_test(mod3)

mod4 <- glmmTMB(cbind(Alive, TotalD) ~ Treatment + Day +
                  (1 | obs_ID),
            family = binomial(), data = dat) 

overdispersion_test(mod4)

AIC(mod1,mod2, mod3, mod4) #mod1 is a better model
```

plotting model
```{r}
plot_model(mod1, type = "pred",
           terms = c("Day [all]", "Treatment"),
           title = "") +
  labs(x = "Day", y = "Pr(Bee Survival)", color = "Treatment") +
  theme_minimal(base_size = 16)

summary(mod1)
car::Anova(mod1)
```


model summary and post-hoc tests
```{r}
summary(mod1)
car::Anova(mod1)

library(emmeans)

emm_day <- emmeans(
  mod1,
  ~ Treatment | Day,
  at = list(Day = c(0:28)),
  type = "response"
)

emm_day
```

reduced dataset
```{r}

datre <- dat %>% 
  filter(dat$Day <= 23)

```

model for datre

``` {r}
mod5 <-  glmmTMB(cbind(Alive, TotalD) ~ Treatment * Day +
                  (1 | CageID) + (1 | obs_ID),
            family = binomial(), data = datre) 

mod6 <-  glmmTMB(cbind(Alive, TotalD) ~ Treatment * Day ,
            family = binomial(), data = datre) 

mod7 <-  glmmTMB(cbind(Alive, TotalD) ~ Treatment * Day +
                  (1 | CageID) ,
            family = binomial(), data = datre) 

mod8 <-  glmmTMB(cbind(Alive, TotalD) ~ Treatment * Day +
                   (1 | obs_ID),
            family = binomial(), data = datre) 

overdispersion_test(mod5) #BEST AIC
overdispersion_test(mod6) #over
overdispersion_test(mod7) #over
overdispersion_test(mod8)
AIC(mod5, mod6, mod7, mod8)

```

plotting model mod5
```{r}
plot_model(mod5, type = "pred",
           terms = c("Day [all]", "Treatment"),
           title = "") +
  labs(x = "Day", y = "Pr(Bee Survival)", color = "Treatment") +
  theme_minimal(base_size = 16)

summary(mod1)
car::Anova(mod1)
```


model summary and post-hoc tests
```{r}
summary(mod5)
car::Anova(mod5)

library(emmeans)

emm_dayre <- emmeans(
  mod5,
  ~ Treatment | Day,
  at = list(Day = c(0:23)),
  type = "response"
)

emm_dayre
```


```{r plots_and_curves, echo = F}

# When is the inflection point?
dat %>% 
  filter(Treatment == "A" | 
           Treatment == "C") %>% 
  ggplot(aes(x = Day, y = logRisk, color = Treatment)) +
  geom_point() +
  geom_smooth(method = "lm",
              formula = y ~ poly(x, 2))
## inflection --- around day 21-23

```

# Bayesian Survival Model

```{r bayes_surv_mods, echo = F}

# get the starting number of bees in each cage:
dat1 <- dat %>% 
  group_by(CageID, Treatment) %>% 
  summarise(total_escape = sum(NBE, na.rm = T)) %>% 
  mutate(total_cens = total_escape + 20) %>% 
  right_join(., dat, by = c("CageID", "Treatment")) %>% 
  mutate(Alive = Alive + total_cens) %>% 
  select(Day, CageID, Treatment, Alive, total_cens, total_escape) %>% 
  group_by(CageID, Treatment) %>% 
  summarise(n_start = max(Alive, na.rm = T))

# calculate the number of at risk bees in each time step:
dat2 <- dat %>% left_join(dat1, by = c("CageID", "Treatment")) %>% 
  mutate(NBE = ifelse(is.na(NBE), 0, NBE)) %>% 
  arrange(CageID, Day) %>% 
  group_by(CageID) %>% 
  mutate(n_atrisk = n_start - cumsum(lag(NDB, default = 0)) -
           cumsum(lag(NBE, default = 0))) %>% 
  ungroup()

# Discrete-time hazard model:
mod1_surv <- brm(
  formula = NDB | trials(n_atrisk) ~ 
    Treatment * splines::ns(Day, df = 3) +
    (1 | CageID),
  family = beta_binomial(link = "logit"),
  data = dat2,
  cores = 4,
  chains = 4,
  prior = c(
    prior(normal(0, 1.5), class = b),
    prior(normal(-3, 1), class = Intercept),
    prior(exponential(1), class = sd)
  ),
  file = "Models/mod1_surv"
)

summary(mod1_surv)

## overdispersion is not really an issue with beta_binomial, but check:
pp_check(mod1_surv, ndraws = 100)
## ^^^^ Observed data (black line) falls in the predicted variance envelope

```

## Visualize the results
```{r predicted_survival_curves_prep, echo = F}

# Build a prediction grid over the observed day range
days <- seq(min(dat2$Day), max(dat2$Day), by = 1)
treatments <- levels(dat2$Treatment)

pred_grid <- expand.grid(Day = days,
                         Treatment = treatments) %>%
  mutate(CageID = NA, # marginal over random effect
         n_atrisk = 1000
         # arbitrary number; we want the probability, not counts
  )

# Posterior draws of the per-day hazard (probability scale)
hazard_draws <- pred_grid %>% 
  add_epred_draws(mod1_surv, re_formula = NA, ndraws = 500) %>% 
  mutate(hazard = .epred / 1000)

```

### Per-day hazard by treatment
```{r hazards_by_treatment, echo = F}

hazard_draws %>% 
  group_by(Day, Treatment) %>% 
  median_qi(hazard, .width = 0.89) %>% 
  ggplot(aes(x = Day, y = hazard,
             color = Treatment,
             fill = Treatment)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper),
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  labs(y = "Per-day probability of death",
       x = "Day") +
  scale_colour_viridis_d(option = "D") +
  scale_fill_viridis_d(option = "D") +
  theme_classic()

```

### Survival curves
```{r survival_curves, echo = F}

survival_draws <- hazard_draws %>% 
  arrange(Treatment, .draw, Day) %>% 
  group_by(Treatment, .draw) %>% 
  mutate(survival = cumprod(1 - hazard)) %>% 
  ungroup()

survival_draws %>% 
  group_by(Day, Treatment) %>% 
  median_qi(survival, .width = 0.89) %>% 
  ggplot(aes(x = Day, y = survival,
             color = Treatment,
             fill = Treatment)) +
  geom_ribbon(aes(ymin = .lower, ymax = .upper),
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::percent) +
  labs(y = "Survival probability", x = "Day") +
  scale_colour_viridis_d(option = "D") +
  scale_fill_viridis_d(option = "D") +
  theme_classic()


```


## Contrast-based hypothesis testing

We'll use the restricted mean survival time, RMST to compare the area under the survival curves across treatments.

```{r RMST_contrasts, echo = F}

days <- sort(unique(dat2$Day))

treatments <- levels(dat2$Treatment)

pred_grid <- expand.grid(Day = days,
                         Treatment = treatments) %>% 
  mutate(CageID = NA, n_atrisk = 1000)

# Get hazard draws
hazard_draws <- pred_grid %>% 
  add_epred_draws(mod1_surv, re_formula = NA, ndraws = 2000) %>% 
  mutate(hazard = .epred / 1000)

# Convert to survival, then compute RMST as sum of survival over days
# (discrete-time RMST = sum of S(t) across all time points)
rmst_draws <- hazard_draws %>% 
  arrange(Treatment, .draw, Day) %>% 
  group_by(Treatment, .draw) %>% 
  summarise(RMST = sum(cumprod(1 - hazard)),
            .groups = "drop")

###################################
# contrasts:

rmst_wide <- rmst_draws %>% 
  pivot_wider(names_from = Treatment, values_from = RMST)

contrasts <- rmst_wide %>% 
  mutate(
    # key hypotheses
    A_vs_C = A - C, # is A better than C?
    A_vs_B = A - B,
    C_vs_B = C - B,
    B_vs_D = B - D,
    B_vs_E = B - E,
    D_vs_E = D - E # are D and E actually distinguishable?
  )

# Summarize each contrast
contrasts %>% 
  select(-A, -B, -C, -D, -E) %>%
  pivot_longer(cols = c(A_vs_C : D_vs_E),
               names_to = "contrast",
               values_to = "diff") %>% 
  group_by(contrast) %>% 
  median_qi(diff, .width = 0.95) %>% 
  mutate(across(where(is.numeric), round, 2))

```

```{r probability_of_direction, echo = F}

contrasts %>% 
  summarise(
    p_A_gt_C = mean(A_vs_C > 0),
    p_A_gt_B = mean(A_vs_B > 0),
    p_C_gt_B = mean(C_vs_B > 0),
    p_B_gt_D = mean(B_vs_D > 0),
    p_B_gt_E = mean(B_vs_E > 0),
    p_D_gt_E = mean(D_vs_E > 0)
  ) %>% 
  pivot_longer(everything(),
               names_to = "contrast",
               values_to = "P(direction)")

```

### Contrast plot
```{r contrast_plot, echo = F}

contrasts %>% 
  pivot_longer(-c(.draw,A,B,C,D,E),
               names_to = "contrast",
               values_to = "diff") %>% 
  ggplot(aes(x = diff, y = contrast)) +
  stat_halfeye(.width = c(0.89, 0.95)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Difference in RMST (days)", y = NULL) +
  theme_classic()


```


